---
title: "Assignment 4 – GCM on the Alien Categorisation Task"
author: "Study Group 5"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(020801)
```

```{r packages}
pacman::p_load(
  tidyverse,
  here,
  cmdstanr,
  posterior,
  bayesplot,
  patchwork
)
```

---

# 1. Stimulus structure

We focus on **Session 1, danger dimension only**.
Rule: an alien is **dangerous** (category 1) if and only if
`eye_stalks == 1 AND spots == 1` (features 1 and 3 in the assignment spec).

```{r stimulus_info}
stimulus_info <- tibble(
  alien_id     = 1:32,
  eye_stalks   = c(1,0,0,0,0,0,1,0,0,0,1,1,1,1,0,0,0,0,0,1,1,1,1,0,0,1,1,1,1,0,1,1),
  colour_green = c(1,0,0,0,0,1,0,0,0,1,0,0,0,1,1,1,0,0,1,0,1,1,1,1,1,0,1,1,1,1,0,0),
  slim_legs    = c(1,0,0,0,1,0,0,0,1,0,0,0,1,0,0,1,1,1,0,0,0,0,1,1,1,1,0,1,1,1,1,1),
  arms_up      = c(1,0,0,1,0,0,0,1,0,0,0,1,0,0,1,0,1,1,1,1,1,0,0,1,1,1,1,0,1,0,1,0),
  spots        = c(1,0,1,0,0,0,0,1,1,1,1,0,0,0,0,0,0,1,1,1,0,1,0,0,1,1,1,1,0,1,0,1),
  # Session 1 danger rule: eye_stalks AND spots
  category_true = as.integer(eye_stalks == 1 & spots == 1)
)

cat("Number of dangerous aliens:", sum(stimulus_info$category_true), "\n")
cat("Number of safe aliens:     ", sum(stimulus_info$category_true == 0), "\n")
```

```{r plot_stimulus, fig.width=14, fig.height=8}
# --- LEFT: scatter with jitter so overlapping binary points separate ---
set.seed(1)
p_structure <- ggplot(stimulus_info,
                      aes(x = eye_stalks, y = spots,
                          colour = factor(category_true))) +
  # jitter slightly so stacked points spread out; binary scale so keep small
  geom_jitter(width = 0.05, height = 0.05, size = 4, alpha = 0.85) +
  # label each point with its alien ID — use a separate geom so label doesn't
  # bleed into the colour legend
  geom_text(aes(label = alien_id),
            position = position_jitter(width = 0.05, height = 0.05, seed = 1),
            nudge_y = 0.07, size = 2.8, colour = "grey20") +
  scale_colour_manual(
    values = c("0" = "#0072B2", "1" = "#D55E00"),
    labels = c("0" = "Safe", "1" = "Dangerous"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = c(0, 1), labels = c("absent", "present")) +
  scale_y_continuous(breaks = c(0, 1), labels = c("absent", "present")) +
  labs(
    title    = "Danger rule: eye_stalks AND spots",
    subtitle = "Each dot = one alien (jittered to avoid overlap)",
    x = "Eye stalks (f1)", y = "Spots (f3)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

# --- RIGHT: heatmap — flip axes so alien ID reads cleanly on x-axis ---
p_heatmap <- stimulus_info %>%
  mutate(
    danger_label = if_else(category_true == 1, "⚠ danger", "safe"),
    alien_label  = paste0(alien_id, " (", danger_label, ")")
  ) %>%
  select(alien_id, alien_label, eye_stalks, colour_green, slim_legs, arms_up, spots) %>%
  pivot_longer(
    cols      = c(eye_stalks, colour_green, slim_legs, arms_up, spots),
    names_to  = "feature",
    values_to = "value"
  ) %>%
  mutate(
    feature = factor(feature,
                     levels = c("eye_stalks","spots","slim_legs","arms_up","colour_green")),
    alien_label = fct_reorder(alien_label, alien_id)
  ) %>%
  ggplot(aes(x = feature, y = alien_label, fill = factor(value))) +
  geom_tile(colour = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = c("0" = "#d9d9d9", "1" = "#fc8d59"),
    labels = c("0" = "absent", "1" = "present"),
    name   = "Feature value"
  ) +
  scale_x_discrete(labels = c(
    eye_stalks   = "Eye stalks\n(f1)",
    spots        = "Spots\n(f3)",
    slim_legs    = "Slim legs\n(f2)",
    arms_up      = "Arms up\n(f4)",
    colour_green = "Green\n(f5)"
  )) +
  labs(
    title    = "All 32 stimuli across 5 features",
    subtitle = "Aliens marked ⚠ are dangerous (eye_stalks AND spots = 1)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y      = element_text(size = 8),
    axis.text.x      = element_text(size = 10),
    legend.position  = "bottom",
    panel.grid       = element_blank()
  )

# Render side by side with independent sizing
p_structure + p_heatmap +
  plot_layout(widths = c(1, 1.4))
```
```{r}
p_heatmap <- stimulus_info %>%
  mutate(
    alien_label  = if_else(category_true == 1,
                           paste0(alien_id),
                           as.character(alien_id)),
    is_dangerous = (eye_stalks == 1 & spots == 1)
  ) %>%
  pivot_longer(
    cols      = c(eye_stalks, colour_green, slim_legs, arms_up, spots),
    names_to  = "feature",
    values_to = "value"
  ) %>%
  mutate(
    fill_type = case_when(
      is_dangerous & feature %in% c("eye_stalks", "spots") ~ "diagnostic",
      value == 1 ~ "present",
      TRUE       ~ "absent"
    ),
    feature     = factor(feature,
                         levels = c("eye_stalks", "spots", "slim_legs", "arms_up", "colour_green"),
                         labels = c("Eye stalks", "Spots", "Slim legs", "Arms up", "Green")),
    # reverse so 1 is at top, 32 at bottom
    alien_label = fct_reorder(factor(alien_label), -alien_id)
  ) %>%
  ggplot(aes(x = feature, y = alien_label, fill = fill_type)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_manual(
    values = c("diagnostic" = "#fc8d59", "present" = "#7bafd4", "absent" = "#e8e8e8"),
    labels = c("diagnostic" = "Eye stalks & Spots = 1 (dangerous)",
               "present"    = "Other feature = 1",
               "absent"     = "Feature = 0"),
    name   = NULL
  ) +
  labs(title = "All 32 stimuli across 5 features", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y      = element_text(size = 9),
    axis.text.x      = element_text(size = 10, face = "bold"),
    legend.position  = "bottom",
    legend.text      = element_text(size = 9),
    panel.grid       = element_blank(),
    plot.title       = element_text(size = 12, face = "bold", margin = margin(b = 8))
  )

# force the plot to be tall enough that 32 rows have room to breathe
ggsave("heatmap_check.png", p_heatmap, width = 8, height = 10)
p_heatmap
```

---

# 2. Helper functions

```{r helpers}
# Attention-weighted city-block distance
distance <- function(vect1, vect2, w) {
  sum(w * abs(vect1 - vect2))
}

# Exponential similarity kernel
similarity <- function(dist_val, c) {
  exp(-c * dist_val)
}
```

---

# 3. Schedule generation

One block = all 32 stimuli in a random order.

```{r make_schedule}
make_subject_schedule <- function(stimulus_info, n_blocks, seed) {
  set.seed(seed)
  n_stim <- nrow(stimulus_info)

  sequence <- unlist(lapply(seq_len(n_blocks), function(b) {
    sample(stimulus_info$alien_id, n_stim, replace = FALSE)
  }))

  tibble(
    trial      = seq_along(sequence),
    block      = rep(seq_len(n_blocks), each = n_stim),
    alien_id   = sequence
  ) %>%
    left_join(stimulus_info, by = "alien_id") %>%
    rename(category_feedback = category_true) %>%
    select(trial, block, alien_id,
           eye_stalks, colour_green, slim_legs, arms_up, spots,
           category_feedback)
}
```

---

# 4. GCM simulator

**Bug fixes from the original file:**

- `nrow(32)` → `nrow(obs)` — `nrow` needs a matrix/data frame, not a number
- `ncol(5)`  → `ncol(obs)` — same issue

Everything else follows section 14.2.4 exactly.

```{r gcm_simulate}
gcm_simulate <- function(w,            # attention weights (sum to 1)
                         c,            # sensitivity (strictly positive)
                         bias,         # response bias towards category 1
                         obs,          # matrix: trials × features
                         cat_feedback, # true labels (0 or 1)
                         quiet = TRUE) {

  # FIX: use the actual dimensions of obs, not hard-coded numbers
  ntrials   <- nrow(obs)
  nfeatures <- ncol(obs)

  prob_cat1    <- numeric(ntrials)
  sim_response <- numeric(ntrials)

  # Start with empty exemplar memory
  memory_obs <- matrix(NA_real_, nrow = 0, ncol = nfeatures)
  memory_cat <- numeric(0)

  for (i in seq_len(ntrials)) {
    if (!quiet && i %% 20 == 0) cat("Trial:", i, "\n")

    current_stim <- as.numeric(obs[i, ])
    n_mem        <- nrow(memory_obs)
    has_cat0     <- any(memory_cat == 0)
    has_cat1     <- any(memory_cat == 1)

    # Cold-start: no exemplars yet, or only one category seen
    if (n_mem == 0 || !has_cat0 || !has_cat1) {
      prob_cat1[i] <- bias

    } else {
      # Compute similarity to every stored exemplar
      sims <- vapply(seq_len(n_mem), function(e) {
        similarity(distance(current_stim, memory_obs[e, ], w), c)
      }, numeric(1))

      # Mean similarity per category (avoids frequency bias)
      m1 <- mean(sims[memory_cat == 1])
      m0 <- mean(sims[memory_cat == 0])

      # Luce choice rule
      num <- bias * m1
      den <- num + (1 - bias) * m0

      prob_cat1[i] <- if (den > 1e-9) num / den else bias
      prob_cat1[i] <- pmax(1e-9, pmin(1 - 1e-9, prob_cat1[i]))
    }

    # Draw binary response
    sim_response[i] <- rbinom(1, 1, prob_cat1[i])

    # Add current stimulus + TRUE feedback label to memory
    memory_obs <- rbind(memory_obs, current_stim)
    memory_cat <- c(memory_cat, cat_feedback[i])
  }

  tibble(prob_cat1 = prob_cat1, sim_response = sim_response)
}
```

---

# 5. Simulate one agent and plot performance

We use all 5 features (with equal attention weights to start) and a moderate sensitivity.

```{r simulate_agent}
# Build schedule for one subject
schedule <- make_subject_schedule(stimulus_info, n_blocks = 1, seed = 42)

obs      <- schedule %>% select(eye_stalks, colour_green, slim_legs, arms_up, spots)
feedback <- schedule$category_feedback

# Parameter choices
w_true    <- rep(1/5, 5)   # equal attention to all 5 features
c_true    <- 2.0           # moderate sensitivity
bias_true <- 0.5           # no response bias

sim_results <- gcm_simulate(
  w = w_true, c = c_true, bias = bias_true,
  obs = obs, cat_feedback = feedback,
  quiet = TRUE
)

# Combine with schedule for plotting
agent_data <- schedule %>%
  mutate(
    prob_cat1    = sim_results$prob_cat1,
    sim_response = sim_results$sim_response,
    correct      = as.integer(category_feedback == sim_response),
    cum_acc      = cumsum(correct) / row_number()
  )
```

```{r plot_performance}
ggplot(agent_data, aes(x = trial, y = cum_acc)) +
  geom_line(colour = "#0072B2", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = sum(feedback == 0) / length(feedback),
             linetype = "dotted", colour = "#D55E00",
             linewidth = 0.8) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "GCM agent: cumulative accuracy over trials",
    subtitle = paste0("w = equal (1/5 each), c = ", c_true, ", bias = ", bias_true,
                      " | orange dotted = majority-class baseline"),
    x = "Trial", y = "Cumulative accuracy"
  ) +
  theme_minimal()
```

```{r plot_prob_cat1}
ggplot(agent_data, aes(x = trial, y = prob_cat1,
                       colour = factor(category_feedback))) +
  geom_point(alpha = 0.7) +
  geom_line(alpha = 0.3, colour = "grey60") +
  scale_colour_manual(values = c("0" = "#0072B2", "1" = "#D55E00"),
                      name = "True category") +
  labs(
    title = "Trial-by-trial P(dangerous) assigned by GCM agent",
    x = "Trial", y = "P(cat 1 = dangerous)"
  ) +
  theme_minimal()
```

---

# 6. Sensitivity analysis: how c affects learning

```{r sensitivity_analysis}
c_values <- c(0.5, 1.0, 2.0, 5.0, 10.0)

sens_results <- map_dfr(c_values, function(cv) {
  res <- gcm_simulate(
    w = w_true, c = cv, bias = bias_true,
    obs = obs, cat_feedback = feedback
  )
  schedule %>%
    mutate(
      c_val   = cv,
      correct = as.integer(category_feedback == res$sim_response),
      cum_acc = cumsum(correct) / row_number()
    )
})

ggplot(sens_results,
       aes(x = trial, y = cum_acc,
           colour = factor(c_val), group = factor(c_val))) +
  geom_line(linewidth = 1) +
  scale_colour_viridis_d(option = "plasma", name = "Sensitivity (c)") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Effect of sensitivity parameter c on cumulative accuracy",
    x = "Trial", y = "Cumulative accuracy"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# 7. Build Stan data list

```{r stan_data}
gcm_stan_data <- list(
  N_total           = nrow(agent_data),
  N_features        = 5L,
  y                 = as.integer(agent_data$sim_response),
  obs               = as.matrix(obs),
  cat_feedback      = as.integer(feedback),
  prior_w_alpha     = rep(1, 5),         # Dirichlet(1,1,1,1,1) – uniform simplex
  prior_log_c_mu    = log(2),            # prior centred on c ≈ 2
  prior_log_c_sigma = 1.0,
  prior_bias_alpha  = 1,
  prior_bias_beta   = 1,
  run_diagnostics   = 1L
)
```

---

# 8. Stan model

Copy-paste of the single-subject GCM from section 14.2.9, adapted to 5 features.

```{r write_stan}
gcm_stan_code <- "
data {
  int<lower=1> N_total;
  int<lower=1> N_features;
  array[N_total] int<lower=0,upper=1> y;
  array[N_total, N_features] real obs;
  array[N_total] int<lower=0,upper=1> cat_feedback;

  vector[N_features] prior_w_alpha;
  real prior_log_c_mu;
  real<lower=0> prior_log_c_sigma;
  real<lower=0> prior_bias_alpha;
  real<lower=0> prior_bias_beta;
  int<lower=0,upper=1> run_diagnostics;
}

transformed data {
  array[N_total, N_total, N_features] real abs_diff;
  for (i in 1:N_total)
    for (j in 1:N_total)
      for (f in 1:N_features)
        abs_diff[i, j, f] = abs(obs[i, f] - obs[j, f]);
}

parameters {
  simplex[N_features] w;
  real log_c;
  real<lower=0,upper=1> bias;
}

transformed parameters {
  real<lower=0> c = exp(log_c);
  vector<lower=1e-9,upper=1-1e-9>[N_total] prob_cat1;

  {
    array[N_total] int memory_trial_idx;
    array[N_total] int memory_cat;
    int n_mem = 0;

    for (i in 1:N_total) {
      real p_i;
      int has_cat0 = 0;
      int has_cat1 = 0;

      for (k in 1:n_mem) {
        if (memory_cat[k] == 0) has_cat0 = 1;
        if (memory_cat[k] == 1) has_cat1 = 1;
      }

      if (n_mem == 0 || has_cat0 == 0 || has_cat1 == 0) {
        p_i = bias;
      } else {
        real s1 = 0; real s0 = 0;
        int  n1 = 0; int  n0 = 0;
        for (e in 1:n_mem) {
          real d = 0;
          int past_i = memory_trial_idx[e];
          for (f in 1:N_features)
            d += w[f] * abs_diff[i, past_i, f];
          real sim = exp(-c * d);
          if (memory_cat[e] == 1) { s1 += sim; n1 += 1; }
          else                    { s0 += sim; n0 += 1; }
        }
        real m1 = s1 / n1;
        real m0 = s0 / n0;
        real num = bias * m1;
        real den = num + (1 - bias) * m0;
        p_i = (den > 1e-9) ? num / den : bias;
      }

      prob_cat1[i] = fmax(1e-9, fmin(1 - 1e-9, p_i));

      n_mem += 1;
      memory_trial_idx[n_mem] = i;
      memory_cat[n_mem] = cat_feedback[i];
    }
  }
}

model {
  target += dirichlet_lpdf(w   | prior_w_alpha);
  target += normal_lpdf(log_c  | prior_log_c_mu, prior_log_c_sigma);
  target += beta_lpdf(bias     | prior_bias_alpha, prior_bias_beta);
  target += bernoulli_lpmf(y   | prob_cat1);
}

generated quantities {
  vector[N_total] log_lik;
  real lprior;
  if (run_diagnostics) {
    for (i in 1:N_total)
      log_lik[i] = bernoulli_lpmf(y[i] | prob_cat1[i]);
  }
  lprior = dirichlet_lpdf(w | prior_w_alpha) +
           normal_lpdf(log_c | prior_log_c_mu, prior_log_c_sigma) +
           beta_lpdf(bias    | prior_bias_alpha, prior_bias_beta);
}
"

stan_file <- here("gcm_alien.stan")
write_stan_file(gcm_stan_code, dir = here(), basename = "gcm_alien.stan")
mod <- cmdstan_model(stan_file)
```

---

# 9. Fit the model to simulated data

```{r fit_model, cache=TRUE}
fit <- mod$sample(
  data            = gcm_stan_data,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 1500,
  adapt_delta     = 0.9,
  refresh         = 300
)
```

---

# 10. Basic diagnostics

```{r diagnostics}
# Key parameters to inspect
params <- c("w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "log_c", "c", "bias")

fit$summary(params)
```

```{r trace_plots}
mcmc_trace(fit$draws(c("log_c", "w[1]", "w[3]", "bias")),
           facet_args = list(ncol = 1))
```

```{r rhat_ess}
mcmc_rhat(rhat(fit))
mcmc_neff(neff_ratio(fit))
```

---

# 11. Prior–Posterior update

```{r prior_posterior}
draws_df <- as_draws_df(fit$draws(c("w[1]","w[3]","log_c","c","bias")))

n_prior <- 4000
prior_df <- tibble(
  `w[1]` = runif(n_prior),
  `w[3]` = runif(n_prior),
  log_c  = rnorm(n_prior, log(2), 1),
  c      = exp(log_c),
  bias   = rbeta(n_prior, 1, 1)
)

pp_panel <- function(prior_x, post_x, true_val, label) {
  tibble(
    x      = c(prior_x, post_x),
    source = c(rep("Prior",     length(prior_x)),
               rep("Posterior", length(post_x)))
  ) %>%
    ggplot(aes(x = x, fill = source, colour = source)) +
    geom_density(alpha = 0.4, linewidth = 0.5) +
    geom_vline(xintercept = true_val, colour = "red",
               linetype = "dashed", linewidth = 0.8) +
    scale_fill_manual(values  = c(Prior = "grey70", Posterior = "#0072B2")) +
    scale_colour_manual(values = c(Prior = "grey50", Posterior = "#0072B2")) +
    labs(title = label, x = NULL, y = "Density") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

p1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, w_true[1],  "w[1] – eye_stalks")
p2 <- pp_panel(prior_df$`w[3]`, draws_df$`w[3]`, w_true[3],  "w[3] – slim_legs")
p3 <- pp_panel(prior_df$log_c,  draws_df$log_c,  log(c_true),"log(c)")
p4 <- pp_panel(prior_df$bias,   draws_df$bias,   bias_true,  "bias")

(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title    = "Prior–Posterior update",
    subtitle = "Red dashed line = true generating value"
  )
```

---

# 12. Posterior predictive check

```{r ppc}
prob_draws <- fit$draws("prob_cat1", format = "matrix")
n_draws    <- nrow(prob_draws)
n_trials   <- ncol(prob_draws)

set.seed(2025)
yrep <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
               nrow = n_draws)

obs_acc <- cummean(agent_data$sim_response == agent_data$category_feedback)
rep_acc <- t(apply(yrep, 1, function(yr) {
  cummean(yr == agent_data$category_feedback)
}))

ppc_summary <- tibble(
  trial = seq_len(n_trials),
  obs   = obs_acc,
  q05   = apply(rep_acc, 2, quantile, 0.05),
  q25   = apply(rep_acc, 2, quantile, 0.25),
  q50   = apply(rep_acc, 2, quantile, 0.50),
  q75   = apply(rep_acc, 2, quantile, 0.75),
  q95   = apply(rep_acc, 2, quantile, 0.95)
)

ggplot(ppc_summary, aes(x = trial)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
  geom_line(aes(y = q50),  colour = "#0072B2", linewidth = 0.8) +
  geom_line(aes(y = obs),  colour = "#D55E00", linewidth = 1.0) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title    = "Posterior Predictive Check",
    subtitle = "Blue ribbons: 50% and 90% PPIs | Orange: observed cumulative accuracy",
    x = "Trial", y = "Cumulative accuracy"
  ) +
  theme_minimal()
```



#Exploring the empirical data
```{r}
# ── Load & prep empirical data ──────────────────────────────────────────────
alien_empirical <- read_csv(here("data", "AlienData.txt")) %>%
  # binary collapse: responses 3/4 = said dangerous, 1/2 = said safe
  mutate(
    resp_danger  = as.integer(response %in% c(3, 4)),
    danger_correct = as.integer(resp_danger == dangerous)
  )

# Session 1, Cycle 1 only — first 32 trials per participant
s1c1 <- alien_empirical %>%
  filter(session == 1, cycle == 1)
```

