---
title: "stan_fittingREAL"
author: "Emma Rose Hahn"
date: "2026-05-11"
output: html_document
---

```{r}
```{r}

pacman::p_load(
  tidyverse, # Data manipulation and plotting
  dplyr,
  here,         # Robust file paths
  cmdstanr, # Stan interface
  ggplot2,     
  posterior,    # Posterior analysis
  bayesplot,
  hrbrthemes, # Plotting MCMC
  plan,
  future, #for the plan function
  furrr,
  patchwork,
  posterior,
  hexbin
)

```


```{r}
# Load the real data
real_data <- read_csv(here("AlienData.txt"), show_col_types = FALSE)

# picking just the  first subject and first session to reduce size (bc there is a LOT)
real_data <- real_data %>%
  filter(subject == 1, session == 1) %>%  # Single subject, single session
  mutate(
    stimulus_code = str_remove(stimulus, "\\.jpg"),
    feature1 = as.numeric(str_sub(stimulus_code, 1, 1)),
    feature2 = as.numeric(str_sub(stimulus_code, 2, 2)),
    feature3 = as.numeric(str_sub(stimulus_code, 3, 3)),
    feature4 = as.numeric(str_sub(stimulus_code, 4, 4)),
    feature5 = as.numeric(str_sub(stimulus_code, 5, 5)),
    y = as.integer(correct),
    cat_feedback = as.integer(correct)
  )

#how much do we have, 208 nrows 
nrow(real_data)

```


```{r}
# Define prior parameters
#LOG_C_PRIOR_MEAN <- log(2)
#LOG_C_PRIOR_SD <- 1


# Prepare real data for Stan
gcm_data_single <- list(
  N_total = nrow(real_data),
  N_features = 5L,
  y = real_data$y,
  obs = as.matrix(real_data[, c("feature1", "feature2", "feature3", "feature4", "feature5")]),
  cat_feedback = real_data$cat_feedback,
  prior_w_alpha = c(1, 1, 1, 1, 1),
  prior_log_c_mu = LOG_C_PRIOR_MEAN,
  prior_log_c_sigma = LOG_C_PRIOR_SD,
  prior_bias_alpha = 1,
  prior_bias_beta = 1,
  run_diagnostics = 1L
)

dir.create(here("simmodels"), showWarnings = FALSE)
fit_filepath_single <- here("simmodels", "stALIEN_fit.rds")

if (regenerate_simulations || !file.exists(fit_filepath_single)) {
  pf_single <- tryCatch(
    mod_gcm_single$pathfinder(
      data      = gcm_data_single,
      seed      = 123,
      num_paths = 4,
      refresh   = 0
    ),
    error = function(e) {
      message("Pathfinder failed — falling back to default initialisation.")
      NULL
    }
  )
  
  fit_gcm_single <- mod_gcm_single$sample(
    data            = gcm_data_single,
    init            = pf_single,
    seed            = 123,
    chains          = 4,
    parallel_chains = min(4, availableCores()),
    iter_warmup     = 1000,
    iter_sampling   = 1500,
    refresh         = 300,
    adapt_delta     = 0.9
  )
  fit_gcm_single$save_object(fit_filepath_single)
  cat("Single-agent model fitted and saved.\n")
} else {
  fit_gcm_single <- readRDS(fit_filepath_single)
  cat("Loaded existing single-agent fit.\n")
}
```

```{r}
diagnostic_summary_table <- function(fit, params) {
  summary_tbl <- fit$summary(params)
  
  # Check diagnostics
  summary_tbl <- summary_tbl %>%
    mutate(
      rhat_pass = rhat < 1.01 | is.na(rhat),
      ess_pass = ess_tail > 400 | is.na(ess_tail),
      pass = rhat_pass & ess_pass
    ) %>%
    select(variable, mean, sd, rhat, ess_bulk, ess_tail, pass)
  
  return(as.data.frame(summary_tbl))  
}

```



```{r}

if (!is.null(fit_gcm_single)) {
  diag_tbl <- diagnostic_summary_table(
  fit_gcm_single,
  params = c("w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "log_c", "c", "bias")
)
  print(diag_tbl)

  if (!all(diag_tbl$pass)) {
    warning("Diagnostic battery has failures — interpret posteriors with caution.")
  }
  

bayesplot::mcmc_trace(
  fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias")),
  facet_args = list(ncol = 1)
) +
  scale_fill_manual(values = c("#FF7518", "#FFBF00", "#6395EE", "#000080")) +
  scale_color_manual(values = c("#FF7518", "#FFBF00", "#6395EE", "#000080"))

bayesplot::mcmc_rank_overlay(
  fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias"))
) +
  scale_fill_manual(values = c("#FF7518", "#FFBF00", "#6395EE", "#000080")) +
  scale_color_manual(values = c("#FF7518", "#FFBF00", "#6395EE", "#000080"))
}
```
```{r}
if (!is.null(fit_gcm_single)) {
  prob_draws <- fit_gcm_single$draws("prob_cat1", format = "matrix")
  n_draws    <- nrow(prob_draws)
  n_trials   <- ncol(prob_draws)
  
  set.seed(020801)
  yrep_post <- matrix(rbinom(n_draws * n_trials, 1, as.vector(prob_draws)),
                      nrow = n_draws)
  
  # Use actual correctness (0/1)
  obs_acc <- cummean(real_data$y)  # Real accuracy
  rep_acc <- t(apply(yrep_post, 1, function(yr) {
    cummean(yr)  # Predicted accuracy
  }))
  
  ppc_summary <- tibble(
    trial = seq_len(n_trials),
    obs   = obs_acc,
    q05   = apply(rep_acc, 2, quantile, 0.05),
    q25   = apply(rep_acc, 2, quantile, 0.25),
    q50   = apply(rep_acc, 2, quantile, 0.50),
    q75   = apply(rep_acc, 2, quantile, 0.75),
    q95   = apply(rep_acc, 2, quantile, 0.95)
  )
  
  ggplot(ppc_summary, aes(x = trial)) +
    geom_ribbon(aes(ymin = q05, ymax = q95), fill = "#56B4E9", alpha = 0.20) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#56B4E9", alpha = 0.40) +
    geom_line(aes(y = q50), color = "#0072B2", linewidth = 0.8) +
    geom_line(aes(y = obs), color = "#D55E00", linewidth = 1) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title    = "Posterior Predictive Check: Cumulative Accuracy",
      subtitle = "Blue ribbons: 50% and 90% posterior predictive intervals\nOrange line: observed cumulative accuracy",
      x = "Trial", y = "Cumulative Accuracy"
    )
}

```

```{r}
if (!is.null(fit_gcm_single)) {
  draws_df <- as_draws_df(
    fit_gcm_single$draws(variables = c("w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "log_c", "c", "bias"))
  )

  set.seed(2026)
  n_prior <- 4000
  prior_df <- tibble(
    `w[1]` = runif(n_prior),
    `w[2]` = runif(n_prior),
    `w[3]` = runif(n_prior),
    `w[4]` = runif(n_prior),
    `w[5]` = runif(n_prior),
    log_c  = rnorm(n_prior, LOG_C_PRIOR_MEAN, LOG_C_PRIOR_SD),
    c      = exp(log_c),
    bias   = rbeta(n_prior, 1, 1)
  )

  pp_panel <- function(prior_x, post_x, true_val, label) {
    tibble(
      x      = c(prior_x, post_x),
      source = c(rep("Prior", length(prior_x)),
                 rep("Posterior", length(post_x)))
    ) |>
      ggplot(aes(x = x, fill = source, color = source)) +
      geom_density(alpha = 0.4, linewidth = 0.5) +
      geom_vline(xintercept = true_val, color = "red",
                 linetype = "dashed", linewidth = 0.7) +
      scale_fill_manual(values = c(Prior = "grey60", Posterior = "#0072B2")) +
      scale_color_manual(values = c(Prior = "grey40", Posterior = "#0072B2")) +
      labs(title = label, x = NULL, y = "Density") +
      theme(legend.position = "bottom")
  }

  p1 <- pp_panel(prior_df$`w[1]`, draws_df$`w[1]`, true_w1, "w[1]")
  p2 <- pp_panel(prior_df$`w[2]`, draws_df$`w[2]`, true_w2, "w[2]")
  p3 <- pp_panel(prior_df$`w[3]`, draws_df$`w[3]`, true_w3, "w[3]")
  p4 <- pp_panel(prior_df$`w[4]`, draws_df$`w[4]`, true_w4, "w[4]")
  p5 <- pp_panel(prior_df$`w[5]`, draws_df$`w[5]`, true_w5, "w[5]")
  p6 <- pp_panel(prior_df$log_c,  draws_df$log_c,  true_log_c, "log(c) — computational scale")
  p7 <- pp_panel(prior_df$c,      draws_df$c,      true_c,     "c — natural scale")
  p8 <- pp_panel(prior_df$bias,   draws_df$bias,   true_bias,  "bias")

  (p1 | p2 | p3) / (p4 | p5 | p6) / (p7 | p8) +
    plot_annotation(
      title    = "Prior–Posterior Update for the Single-Subject GCM",
      subtitle = "Red dashed line = true generating value"
    )
}

```
```{r}
bayesplot::mcmc_pairs(
  fit_gcm_single$draws(c("log_c", "w[1]", "w[2]", "w[3]", "w[4]", "w[5]", "bias")),
  diag_fun     = "dens",
  off_diag_fun = "hex"
)

```

